---
title: "75 max transcript enrichment"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  # github_document:
  #   toc: true
  #   toc_depth: 3
  #   number_sections: true
  #   html_preview: true
---

In this document, I'll take information about changes in the maximum number of transcripts expressed and investigate 1) the significance of these changes, 2) functions associated with these changes and 3) determine if there are any overrepresented biological processes associated with changes

# Set up R Markdown document

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = TRUE,     # Hide warnings
  message = TRUE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)

knitr::opts_knit$set(root.dir = normalizePath("../output/75-max-transcript-enrichment/")) #Set root directory
```

```{r}
getwd()
```

# Install packages

```{r setup, include=FALSE}
# if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')
# if ("ggpubr" %in% rownames(installed.packages()) == 'FALSE') install.packages('ggpubr')
# if ("data.table" %in% rownames(installed.packages()) == 'FALSE') install.packages('data.table')
# if ("RColorBrewer" %in% rownames(installed.packages()) == 'FALSE') install.packages('RColorBrewer')
# if ("patchwork" %in% rownames(installed.packages()) == 'FALSE') install.packages('patchwork')
```

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("topGO")
```

```{r}
require(tidyverse)
require(ggpubr)
require(data.table)
require(RColorBrewer)
require(patchwork)
require(topGO)
```

```{r}
sessionInfo()
```

# Import data

```{r}
metadata <- read.csv("../../data/adult-meta.csv", header = TRUE) %>%
  dplyr::rename(., sample = Sample.ID) %>%
  dplyr::rename(., treatment = Treatment) %>%
  mutate(., treatment = gsub(treatment, pattern = "Control", replacement = "control")) %>%
  mutate(., treatment = gsub(treatment, pattern = "Exposed", replacement = "exposed")) #Import metadata. Change column name and contents to be consistent throughout script
head(metadata)
```

## Maximum transcript data

```{r}
female_tr <- read.csv("../../supplemental-files/01.01-fmcoe-max-predom-isos-gene_fpkm.csv") %>%
  dplyr::select(., c(1:2,4)) %>%
  mutate(., difference = females_exposed_max_transcript_counts - females_controls_max_transcript_counts) #Import transcript count data per gene. Select only the data for females. Create a difference column for exposed vs. control
nrow(female_tr)
head(female_tr)
```

```{r}
male_tr <- read.csv("../../supplemental-files/01.01-fmcoe-max-predom-isos-gene_fpkm.csv") %>%
  dplyr::select(., c(1,3,5)) %>%
  mutate(., difference = males_exposed_max_transcript_counts - males_controls_max_transcript_counts) #Import transcript count data per gene. Select only the data for females. Create a difference column for exposed vs. control
head(male_tr) #Confirm import
```

## Methylation data

```{r}
geneMethylation <- read.csv("../40-gene-methylaiton.csv", header = TRUE, row.names = 2)
geneMethylationMod <- geneMethylation %>%
  dplyr::select(., -1) #Remove extra column
head(geneMethylationMod)
```

```{r}
geneMethylationFem <- geneMethylationMod %>%
  filter(., grepl("F", rownames(.), fixed = TRUE)) %>% 
  t(.) %>%
  as.data.frame(.) %>%
  drop_na(.) %>%
  as.data.frame(.) %>% 
  rownames_to_column(., var = "gene_name") %>%
  pivot_longer(., !gene_name, names_to = "sample", values_to = "meth") %>%
  mutate(., gene_name = gsub(pattern = "gene.", replacement = "", gene_name)) %>%
  left_join(., metadata, by = "sample") %>%
  dplyr::select(., c(gene_name, meth, treatment)) %>%
  dplyr::group_by(gene_name, treatment) %>%
  summarise(avg_meth = mean(meth, na.rm = TRUE)) %>%
  ungroup(.) #Filter female samples and reformat data. Remove excess information from gene name column. Add treatment information and average methylation for each gene by treatment.
head(geneMethylationFem) #Confirm formatting. Resultant df has one line per gene/treatment.
```

```{r}
geneMethylationFemWide <- geneMethylationFem %>%
  pivot_wider(., names_from = treatment, names_prefix = "avg_meth_", values_from = avg_meth) %>%
  replace(is.na(.), 0) %>%
  mutate(., change_meth = avg_meth_exposed - avg_meth_control) #Pivot dataaframe wider and calculate change in gene methylation
head(geneMethylationFemWide)
```

```{r}
geneMethylationMale <- geneMethylationMod %>%
  filter(., grepl("M", rownames(.), fixed = TRUE)) %>% 
  t(.) %>%
  as.data.frame(.) %>%
  drop_na(.) %>%
  as.data.frame(.) %>% 
  rownames_to_column(., var = "gene_name")  %>%
  pivot_longer(., !gene_name, names_to = "sample", values_to = "meth") %>%
  mutate(., gene_name = gsub(pattern = "gene.", replacement = "", gene_name)) %>%
  left_join(., metadata, by = "sample") %>%
  dplyr::select(., c(gene_name, meth, treatment)) %>%
  dplyr::group_by(gene_name, treatment) %>%
  summarise(avg_meth = mean(meth, na.rm = TRUE)) %>%
  ungroup(.) #Filter male samples and reformat data. Remove excess information from gene name column. Add treatment information and average methylation for each gene by treatment.
head(geneMethylationMale) #Confirm formatting
```

```{r}
geneMethylationMaleWide <- geneMethylationMale %>%
  pivot_wider(., names_from = treatment, names_prefix = "avg_meth_", values_from = avg_meth) %>%
  replace(is.na(.), 0) %>%
  mutate(., change_meth = avg_meth_exposed - avg_meth_control) #Pivot dataaframe wider and calculate change in gene methylation
head(geneMethylationMaleWide)
```

```{r}
DMLLocationsFemale <- read.delim("../DML-characterization/female_dml-Gene-wb.bed", sep = "\t", header = FALSE, col.names = c("chr", "start", "end", "f_DML", "meth.diff", "gene.chr", "Gnomon", "gene", "gene.start", "gene.end", "V11", "strand", "V13", "V14")) %>%
  dplyr::select(., "chr", "start", "end", "meth.diff", "gene.start", "gene.end", "V14") %>%
  separate(., col = V14, into = c("gene_name", "misc"), sep = ";Dbxref=GeneID:") %>%
  dplyr::select(., -misc) %>%
  mutate(., gene_name = gsub(pattern = "ID=gene-", replacement = "", x = gene_name)) #Import bedfile with DML locations in genes with gene ID annotations. Rename columns as file is imported then select columns of interest. Separate gene ID annotation information from misc information, and remove misc column. Remove extra information before gene ID
head(DMLLocationsFemale) #Confirm file import and changes
```

```{r}
DMLLocationsMale <- read.delim("../DML-characterization/male_dml-Gene-wb.bed", sep = "\t", header = FALSE, col.names = c("chr", "start", "end", "f_DML", "meth.diff", "gene.chr", "Gnomon", "gene", "gene.start", "gene.end", "V11", "strand", "V13", "V14")) %>%
  dplyr::select(., "chr", "start", "end", "meth.diff", "gene.start", "gene.end", "V14") %>%
  separate(., col = V14, into = c("gene_name", "misc"), sep = ";Dbxref=GeneID:") %>%
  dplyr::select(., -misc) %>%
  mutate(., gene_name = gsub(pattern = "ID=gene-", replacement = "", x = gene_name)) #Import bedfile with DML locations in genes with gene ID annotations. Rename columns as file is imported then select columns of interest. Separate gene ID annotation information from misc information, and remove misc column. Remove extra information before gene ID
head(DMLLocationsMale) #Confirm file import and changes
```

## Expression data

```{r}
geneExpression <- read.csv("../../data/gene_fpkm.csv", header = TRUE, row.names = metadata$sample)
geneExpressionMod <- geneExpression %>%
  dplyr::select(., -c(1:6)) #Remove extra information to have just FPKM for each gene
head(geneExpressionMod)
```

```{r}
geneExpressionFem <- geneExpressionMod %>%
  filter(., grepl("F", rownames(.), fixed = TRUE)) %>%
  t(.) %>%
  as.data.frame(.) %>%
  drop_na(.) %>%
  as.data.frame(.) %>% 
  rownames_to_column(., var = "gene_name")  %>%
  pivot_longer(., cols = !gene_name, names_to = "sample", values_to = "exp") %>% 
  mutate(., gene_name = gsub("gene.", "", gene_name))  %>%
  left_join(., metadata, by = "sample") %>%
  dplyr::select(., c(gene_name, exp, treatment)) %>%
  dplyr::group_by(gene_name, treatment) %>%
  summarise(avg_exp = mean(exp, na.rm = TRUE)) %>%
  ungroup(.) #Filter female samples and reformat data. Remove excess information from gene name column. Add treatment information and average expression for each gene by treatment.
head(geneExpressionFem) #Confirm formatting
```

```{r}
geneExpressionFem %>%
  filter(., avg_exp != 0) %>%
  dplyr::select(., gene_name) %>%
  distinct(.) %>% nrow(.)
```

```{r}
geneExpressionFemWide <- geneExpressionFem %>%
  pivot_wider(., names_from = treatment, names_prefix = "avg_exp_", values_from = avg_exp) %>%
  replace(is.na(.), 0) %>%
  mutate(., change_exp = avg_exp_exposed - avg_exp_control) #Pivot dataaframe wider and calculate change in gene Expression
head(geneExpressionFemWide)
```

```{r}
geneExpressionMale <- geneExpressionMod %>%
  filter(., grepl("M", rownames(.), fixed = TRUE)) %>%
  t(.) %>%
  as.data.frame(.) %>%
  drop_na(.) %>%
  as.data.frame(.) %>% 
  rownames_to_column(., var = "gene_name")  %>%
  pivot_longer(., !gene_name, names_to = "sample", values_to = "exp") %>% 
  mutate(., gene_name = gsub("gene.", "", gene_name)) %>%
  left_join(., metadata, by = "sample") %>%
  dplyr::select(., c(gene_name, exp, treatment)) %>%
  dplyr::group_by(gene_name, treatment) %>%
  summarise(avg_exp = mean(exp, na.rm = TRUE)) %>%
  ungroup(.) #Filter male samples and reformat data. Remove excess information from gene name column. Add treatment information and average expression for each gene by treatment.
head(geneExpressionMale) #Confirm formatting
```

```{r}
geneExpressionMale %>%
  filter(., avg_exp != 0) %>%
  dplyr::select(., gene_name) %>%
  unique(.) %>% nrow(.)
```

```{r}
geneExpressionMaleWide <- geneExpressionMale %>%
  pivot_wider(., names_from = treatment, names_prefix = "avg_exp_", values_from = avg_exp) %>%
  replace(is.na(.), 0) %>%
  mutate(., change_exp = avg_exp_exposed - avg_exp_control) #Pivot dataaframe wider and calculate change in gene Expression
head(geneExpressionMaleWide)
```

## Annotation data

```{r}
mRNATrack <- read.delim("../../genome-features/C_virginica-3.0-mRNA.gff", header = FALSE) %>%
  separate(., V9, into = c("V10", "V11"), sep = "ID=rna-") %>%
  separate(., V11, into = c("V12", "V13"), sep = ";Parent=") %>%
  separate(., V13, into = c("V14", "V15"), sep = ";Dbxref=") %>%
  separate(., V15, into = c("V16", "V17"), sep = ";product=") %>%
  separate(., V17, into = c("V18", "V19"), sep = ";transcript_id=") %>%
  dplyr::select(., c("V12", "V14", "V18")) %>%
  dplyr::rename(., transcript = V12, gene = V14, product = V18) #Winnow down mRNA track
mRNATrack$gene_name <- gsub("gene-", "", mRNATrack$gene) #Make gene naming consistent
head(mRNATrack) #Confirm changes
```

```{r}
GOterms <- read.delim("../../genome-features/_blast-GO-unfolded.sorted", header = FALSE, col.names = c("GO", "transcript"))
head(GOterms)
```

```{r}
GOslim <- read.delim("../../genome-features/GO-GOslim.sorted", header = FALSE, col.names = c("GO", "GOterm", "GOslim", "process"))
head(GOslim)
```

# Females

## Maximum transcript change

```{r}
nrow(female_tr) #39104 genes total
sum(female_tr$difference > 0) #1529 genes with more transcripts expressed in exposed
sum(female_tr$difference == 0) #35048
sum(female_tr$difference < 0) #2437 genes with less transcripts expressed in exposed
sum(female_tr$difference > 0) / nrow(female_tr) * 100 #3.92% of genes have more transcripts expressed in exposed
sum(female_tr$difference == 0) / nrow(female_tr) * 100 #89.83% of genes have more transcripts expressed in exposed
sum(female_tr$difference < 0) / nrow(female_tr) * 100 #6.25% of genes have less transcripts expressed in exposed
```

```{r}
maxTransIncFemTest <- prop.test(x = 1529, n = 39104, p = 0.33,
                                alternative = "two.sided", 
                                conf.level = 0.95) #Conduct a 1-sample z test
maxTransIncFemTest
maxTransIncFemTest$statistic #X-squared = 14965
maxTransIncFemTest$p.value #p = 0. Increases in max transcripts happen significantly less than what would be expected by random chance
```

```{r}
maxTransDecFemTest <- prop.test(x = 2437, n = 39104, p = 0.33,
                                alternative = "two.sided", 
                                conf.level = 0.95) #Conduct a 1-sample z test
maxTransDecFemTest
maxTransDecFemTest$statistic #X-squared = 12671.25
maxTransDecFemTest$p.value #p = 0. Decreases in max transcripts happen significantly less than what would be expected by random chance
```

## Annotate data

```{r}
maxTransFemAnnot <- inner_join(female_tr, mRNATrack, by = "gene_name") %>%
  inner_join(., GOterms, by = "transcript") %>%
  inner_join(., GOslim, by = "GO")
head(maxTransFemAnnot)
```

## Enrichment test

I will use `topGO` to understand if there are biological processes overrepresented in the list of genes that had changes in the maximum number of transcripts expressed when compared to all genes with expression data. I will examine genes with an increase or decrease in maximum transcripts in exposed samples separate for enrichment.

### Make input files

First thing I need to do is modify the GOterms list so each transcript ID has all matching GOterms in one line. I then need to pull out only the genes with data for > 1 transcript, since this will be the background list

```{r}
maxTransFemAnnot %>%
  dplyr::select(., gene_name, GO) %>%
  unique(.) %>%
  group_by(., gene_name) %>%
  summarise(GO = paste(GO, collapse = ",")) %>%
  write.table(., "geneid2go-fem_maxTrans.tab", quote = FALSE, sep = "\t", col.names = FALSE, row.names = FALSE)
#Take list of genes and GOterms for female maximum transcript data, and remove any duplicate entries (precaution). Using genes since that was the "input" for the maximum transcrpt analysis. Group by gene ID, then summarise using paste to get all GOterms for each transcript into the same row separated by ",". Save as a tab delimited file
```

```{r}
geneID2GO <- readMappings(file = "geneid2go-fem_maxTrans.tab") #Loading the GO annotations and GeneIDs. Each line has one transcript ID and all associated GOterms
str(head(geneID2GO)) #Confirm file structure
length(geneID2GO) #31,501 genes with annotations out of 39,104 genes with data for ≥ 2 transcripts
```

```{r}
geneNames <- names(geneID2GO) #Extract names to use as a gene universe
head(geneNames)
```

```{r}
maxTransFemPos <- female_tr %>%
  filter(., difference > 0) %>%
  dplyr::select(., gene_name) #Take data and retain genes where the predominant isoform shifted (difference > 0). Select the gene ID column and convert to a dataframe. Save as a new object
maxTransFemPos <- maxTransFemPos$gene_name #Save column as a vector (will not work otherwise!)
```

```{r}
maxTransFemPosGeneList <- factor(as.integer(geneNames %in% maxTransFemPos))  #Create a factor vector to indicate genes that increased transcripts as significant (1) and didn't as not significant (0)
names(maxTransFemPosGeneList) <- geneNames
str(maxTransFemPosGeneList)
```

```{r}
maxTransFemNeg <- female_tr %>%
  filter(., difference < 0) %>%
  dplyr::select(., gene_name) #Take data and retain genes where the predominant isoform shifted (difference < 0). Select the gene ID column and convert to a dataframe. Save as a new object
maxTransFemNeg <- maxTransFemNeg$gene_name #Save column as a vector (will not work otherwise!)
```

```{r}
maxTransFemNegGeneList <- factor(as.integer(geneNames %in% maxTransFemNeg))  #Create a factor vector to indicate genes that increased transcripts as significant (1) and didn't as not significant (0)
names(maxTransFemNegGeneList) <- geneNames
str(maxTransFemNegGeneList)
```

### Increases in maximum transcripts

#### Run enrichment test

I am only going to look at biological processes.

```{r}
maxTransFemPosDataBP <- new("topGOdata", ontology = "BP", allGenes = maxTransFemPosGeneList,
                            annot = annFUN.gene2GO, gene2GO = geneID2GO) #Create biological process topGO object
maxTransFemPosDataBP #Get summary of object
```

```{r}
test.stat <- new("classicCount", testStatistic = GOFisherTest, name = "Fisher test")
resultFisher.maxTransFemPosDataBP <- getSigGroups(maxTransFemPosDataBP, test.stat)
resultFisher.maxTransFemPosDataBP
```

```{r}
pvalFis.maxTransFemPosDataBP <- score(resultFisher.maxTransFemPosDataBP) #Extract p-values
head(pvalFis.maxTransFemPosDataBP)
hist(pvalFis.maxTransFemPosDataBP, 50, xlab = "p-values") #Plot histogram of p-values
```

```{r}
allRes.maxTransFemPosDataBP <- GenTable(maxTransFemPosDataBP, classic = resultFisher.maxTransFemPosDataBP, ranksOf = "classic", orderBy = "classic", topNodes = length(pvalFis.maxTransFemPosDataBP))
#Create a statistical results table with statistical test results. Order by p-value (classic), and include all results (topNodes)
head(allRes.maxTransFemPosDataBP)
```

```{r}
write.csv(allRes.maxTransFemPosDataBP, "fem-maxTransPos-BP-FisherTestResults.csv", quote = FALSE, row.names = FALSE) #Save dataframe
```

#### Match enriched GOterms with general annotation information

```{r}
sigRes.maxTransFemPosDataBP <- allRes.maxTransFemPosDataBP[1:211,c(1, 6)] #Filter significantly enriched GOterms, only keep GOID and p-value
colnames(sigRes.maxTransFemPosDataBP) <- c("GO", "p.value") #Change column names
head(sigRes.maxTransFemPosDataBP)
```

```{r}
sigRes.maxTransFemPosDataBP %>%
  left_join(., y = maxTransFemAnnot, by = "GO") %>%
  filter(., difference > 0) %>%
  dplyr::select(., gene_name, difference, females_controls_max_transcript_counts, females_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  write.csv("fem-maxTransPos-BP-EnrichedGO-geneAnnot.csv", quote = FALSE, row.names = FALSE) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Save as a csv file.
```

### Decreases in maximum transcripts

#### Run enrichment test

```{r}
maxTransFemNegDataBP <- new("topGOdata", ontology = "BP", allGenes = maxTransFemNegGeneList,
                            annot = annFUN.gene2GO, gene2GO = geneID2GO) #Create biological process topGO object
maxTransFemNegDataBP #Get summary of object
```

```{r}
test.stat <- new("classicCount", testStatistic = GOFisherTest, name = "Fisher test")
resultFisher.maxTransFemNegDataBP <- getSigGroups(maxTransFemNegDataBP, test.stat)
resultFisher.maxTransFemNegDataBP
```

```{r}
pvalFis.maxTransFemNegDataBP <- score(resultFisher.maxTransFemNegDataBP) #Extract p-values
head(pvalFis.maxTransFemNegDataBP)
hist(pvalFis.maxTransFemNegDataBP, 50, xlab = "p-values") #Plot histogram of p-values
```

```{r}
allRes.maxTransFemNegDataBP <- GenTable(maxTransFemNegDataBP, classic = resultFisher.maxTransFemNegDataBP, ranksOf = "classic", orderBy = "classic", topNodes = length(pvalFis.maxTransFemNegDataBP))
#Create a statistical results table with statistical test results. Order by p-value (classic), and include all results (topNodes)
head(allRes.maxTransFemNegDataBP)
```

```{r}
write.csv(allRes.maxTransFemNegDataBP, "fem-maxTransNeg-BP-FisherTestResults.csv", quote = FALSE, row.names = FALSE) #Save dataframe
```

#### Match enriched GOterms with general annotation information

```{r}
sigRes.maxTransFemNegDataBP <- allRes.maxTransFemNegDataBP[1:469,c(1, 6)] #Filter significantly enriched GOterms, only keep GOID and p-value
colnames(sigRes.maxTransFemNegDataBP) <- c("GO", "p.value") #Change column names
head(sigRes.maxTransFemNegDataBP)
```

```{r}
sigRes.maxTransFemNegDataBP %>%
  left_join(., y = maxTransFemAnnot, by = "GO") %>%
  filter(., difference < 0) %>%
  dplyr::select(., gene_name, difference, females_controls_max_transcript_counts, females_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  write.csv("fem-maxTransNeg-BP-EnrichedGO-geneAnnot.csv", quote = FALSE, row.names = FALSE) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Save as a csv file.
```

# Changes in maximum transcripts expressed and methylation

## Format data

```{r}
maxTransFemWide <- inner_join(female_tr, geneMethylationFemWide, by = "gene_name") %>%
  inner_join(., geneExpressionFemWide, by = "gene_name") #Join maximum transcript expression data with gene methylation and expression data
head(maxTransFemWide)
```

## Methylation of genes with enriched BPGO

### Increases in maximum transcripts

```{r}
maxTransFemPosAnnot <- sigRes.maxTransFemPosDataBP %>%
  left_join(., y = maxTransFemAnnot, by = "GO") %>%
  dplyr::select(., gene_name, difference, females_controls_max_transcript_counts, females_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  full_join(x = maxTransFemWide, y = ., by = "gene_name") %>%
  dplyr::select(., gene_name, difference.x, change_meth, change_exp, GO, p.value, product) %>%
  dplyr::rename(., difference = "difference.x") %>% 
  filter(., difference > 0)
#Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Full join with the wide maximum transcript dataframe that has change in methylation and expression data. Select relevant columns and rename change column.
head(maxTransFemPosAnnot) #Confirm manipulation
nrow(maxTransFemPosAnnot)
```

```{r}
femMaxTransPosEnrichedBPGO <- maxTransFemPosAnnot %>%
  filter(., is.na(GO) == FALSE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "", y = "Number of Genes", title = "A. Enriched BPGO") +
  theme_classic(base_size = 15)

femMaxTransPosNotEnrichedBPGO <- maxTransFemPosAnnot %>%
  filter(., is.na(GO) == TRUE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "Change in Methylation (%)", y = "Number of Genes", title = "B. Not Enriched BPGO") +
  theme_classic(base_size = 15)
```

```{r}
femMaxTransPosEnrichedBPGO / femMaxTransPosNotEnrichedBPGO
ggsave("fem-max-transcripts-positive-changeMeth-enrichedBPGO.pdf", height = 8.5, width = 11)
```

### Decreases in maximum transcripts

```{r}
maxTransFemNegAnnot <- sigRes.maxTransFemNegDataBP %>%
  left_join(., y = maxTransFemAnnot, by = "GO") %>%
  dplyr::select(., gene_name, difference, females_controls_max_transcript_counts, females_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  full_join(x = maxTransFemWide, y = ., by = "gene_name") %>%
  dplyr::select(., gene_name, difference.x, change_meth, change_exp, GO, p.value, product) %>%
  dplyr::rename(., difference = "difference.x") %>%
  filter(., difference < 0) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Full join with the wide maximum transcript dataframe that has change in methylation and expression data. Select relevant columns and rename change column.
head(maxTransFemNegAnnot) #Confirm manipulation
```

```{r}
femMaxTransNegEnrichedBPGO <- maxTransFemNegAnnot %>%
  filter(., is.na(GO) == FALSE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "", y = "Number of Genes", title = "A. Enriched BPGO") +
  theme_classic(base_size = 15)

femMaxTransNegNotEnrichedBPGO <- maxTransFemNegAnnot %>%
  filter(., is.na(GO) == TRUE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "Change in Methylation (%)", y = "Number of Genes", title = "B. Not Enriched BPGO") +
  theme_classic(base_size = 15)
```

```{r}
femMaxTransNegEnrichedBPGO / femMaxTransNegNotEnrichedBPGO
ggsave("fem-max-transcripts-negative-changeMeth-enrichedBPGO.pdf", height = 8.5, width = 11)
```

## Number of DML in genes with changes in the maximum transcripts expressed

```{r}
maxTransFemWideDML <- left_join(maxTransFemWide, DMLLocationsFemale, by = "gene_name") %>%
  dplyr::select(., -c(chr, gene.start, gene.end)) %>%
  group_by(., gene_name) %>%
  mutate(., DMLcount = n()) %>%
  mutate(., DMLcount = case_when(is.na(meth.diff) == TRUE ~ 0,
                                 is.na(meth.diff) == FALSE ~ DMLcount)) #Join max trans data with DML annotations and remove superfluous columns. Group by gene name and count the number of DML per gene. When there is no DML in the gene (meth.diff == NA), change the DML count to 0.
head(maxTransFemWideDML) #Confirm dataframe creation
```

```{r}
write.csv(maxTransFemWideDML, "fem-max-trans-DML-annotations.csv", quote = FALSE, row.names = FALSE) #Save dataframe
```

```{r}
maxTransFemWideDML %>%
  group_by(., DMLcount, difference) %>%
  summarise(., count = n()) #Group data by DML count and change, then count the number of genes with or without predominant isoform shifts with various DML counts
```

DMLcount difference count
0	       -17	      1		
0	       -16	      1		
0	       -15      	1		
0	       -13	      1		
0	       -10	      2		
0	       -9	        3		
0	       -8	        1		
0	       -7	        3		
0	       -6	        5		
0	       -5	        10
0	       -4	        26		
0	       -3	        72		
0	       -2	        273		
0	       -1	        1624		
0        0	        29283		
0	       1	        1067		
0	       2	        106		
0        3	        15		
0	       4	        5		
0	       5	        1	
0	       7	        1		
1	       -3	        1		
1	       0	        53		
1	       1	        1		
2	       -2	        4		
2	       -1	        2		
2	       0	        10		
3	       0	        3	

# Males

## Maximum transcript change

```{r}
nrow(male_tr) #39104 genes total
sum(male_tr$difference > 0) #3123 genes with more transcripts expressed in exposed
sum(male_tr$difference == 0) #34687
sum(male_tr$difference < 0) #1204 genes with less transcripts expressed in exposed
sum(male_tr$difference > 0) / nrow(male_tr) * 100 #8.00% of genes have more transcripts expressed in exposed
sum(male_tr$difference == 0) / nrow(male_tr) * 100 #88.9% of genes have more transcripts expressed in exposed
sum(male_tr$difference < 0) / nrow(male_tr) * 100 #3.09% of genes have less transcripts expressed in exposed
```

```{r}
maxTransIncMaleTest <- prop.test(x = 3123, n = 39104, p = 0.33,
                                 alternative = "two.sided", 
                                 conf.level = 0.95) #Conduct a 1-sample z test
maxTransIncMaleTest
maxTransIncMaleTest$statistic #X-squared = 11064.72
maxTransIncMaleTest$p.value #p = 0. Increases in max transcripts happen significantly less than what would be expected by random chance
```

```{r}
maxTransDecMaleTest <- prop.test(x = 1204, n = 39104, p = 0.33,
                                 alternative = "two.sided", 
                                 conf.level = 0.95) #Conduct a 1-sample z test
maxTransDecMaleTest
maxTransDecMaleTest$statistic #X-squared = 15832.46
maxTransDecMaleTest$p.value #p = 0. Decreases in max transcripts happen significantly less than what would be expected by random chance
```

## Annotate data

```{r}
maxTransMaleAnnot <- inner_join(male_tr, mRNATrack, by = "gene_name") %>%
  inner_join(., GOterms, by = "transcript") %>%
  inner_join(., GOslim, by = "GO")
head(maxTransMaleAnnot)
```

## Enrichment test

## Make input files

```{r}
maxTransMaleAnnot %>%
  dplyr::select(., gene_name, GO) %>%
  unique(.) %>%
  group_by(., gene_name) %>%
  summarise(GO = paste(GO, collapse = ",")) %>%
  write.table(., "geneid2go-male_maxTrans.tab", quote = FALSE, sep = "\t", col.names = FALSE, row.names = FALSE)
#Take list of genes and GOterms for male maximum transcript data, and remove any duplicate entries (precaution). Using genes since that was the "input" for the maximum transcrpt analysis. Group by gene ID, then summarise using paste to get all GOterms for each transcript into the same row separated by ",". Save as a tab delimited file
```

```{r}
geneID2GO <- readMappings(file = "geneid2go-male_maxTrans.tab") #Loading the GO annotations and GeneIDs. Each line has one transcript ID and all associated GOterms
str(head(geneID2GO)) #Confirm file structure
length(geneID2GO) #31,501 genes with annotations out of 39,104 genes with data for ≥ 2 transcripts
```

```{r}
geneNames <- names(geneID2GO) #Extract names to use as a gene universe
head(geneNames)
```

```{r}
maxTransMalePos <- male_tr %>%
  filter(., difference > 0) %>%
  dplyr::select(., gene_name) #Take data and retain genes where the predominant isoform shifted (difference > 0). Select the gene ID column and convert to a dataframe. Save as a new object
maxTransMalePos <- maxTransMalePos$gene_name #Save column as a vector (will not work otherwise!)
```

```{r}
maxTransMalePosGeneList <- factor(as.integer(geneNames %in% maxTransMalePos))  #Create a factor vector to indicate genes that increased transcripts as significant (1) and didn't as not significant (0)
names(maxTransMalePosGeneList) <- geneNames
str(maxTransMalePosGeneList)
```

```{r}
maxTransMaleNeg <- male_tr %>%
  filter(., difference < 0) %>%
  dplyr::select(., gene_name) #Take data and retain genes where the predominant isoform shifted (difference < 0). Select the gene ID column and convert to a dataframe. Save as a new object
maxTransMaleNeg <- maxTransMaleNeg$gene_name #Save column as a vector (will not work otherwise!)
```

```{r}
maxTransMaleNegGeneList <- factor(as.integer(geneNames %in% maxTransMaleNeg))  #Create a factor vector to indicate genes that increased transcripts as significant (1) and didn't as not significant (0)
names(maxTransMaleNegGeneList) <- geneNames
str(maxTransMaleNegGeneList)
```

### Increases in maximum transcripts

#### Run enrichment test

```{r}
maxTransMalePosDataBP <- new("topGOdata", ontology = "BP", allGenes = maxTransMalePosGeneList,
                             annot = annFUN.gene2GO, gene2GO = geneID2GO) #Create biological process topGO object
maxTransMalePosDataBP #Get summary of object
```

```{r}
test.stat <- new("classicCount", testStatistic = GOFisherTest, name = "Fisher test")
resultFisher.maxTransMalePosDataBP <- getSigGroups(maxTransMalePosDataBP, test.stat)
resultFisher.maxTransMalePosDataBP
```

```{r}
pvalFis.maxTransMalePosDataBP <- score(resultFisher.maxTransMalePosDataBP) #Extract p-values
head(pvalFis.maxTransMalePosDataBP)
hist(pvalFis.maxTransMalePosDataBP, 50, xlab = "p-values") #Plot histogram of p-values
```

```{r}
allRes.maxTransMalePosDataBP <- GenTable(maxTransMalePosDataBP, classic = resultFisher.maxTransMalePosDataBP, ranksOf = "classic", orderBy = "classic", topNodes = length(pvalFis.maxTransMalePosDataBP))
#Create a statistical results table with statistical test results. Order by p-value (classic), and include all results (topNodes)
head(allRes.maxTransMalePosDataBP)
```

```{r}
write.csv(allRes.maxTransMalePosDataBP, "male-maxTransPos-BP-FisherTestResults.csv", quote = FALSE, row.names = FALSE) #Save dataframe
```

#### Match enriched GOterms with general annotation information

```{r}
sigRes.maxTransMalePosDataBP <- allRes.maxTransMalePosDataBP[1:391,c(1, 6)] #Filter significantly enriched GOterms, only keep GOID and p-value
colnames(sigRes.maxTransMalePosDataBP) <- c("GO", "p.value") #Change column names
head(sigRes.maxTransMalePosDataBP)
```

```{r}
sigRes.maxTransMalePosDataBP %>%
  left_join(., y = maxTransMaleAnnot, by = "GO") %>%
  filter(., difference > 0) %>%
  dplyr::select(., gene_name, difference, males_controls_max_transcript_counts, males_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  write.csv("male-maxTransPos-BP-EnrichedGO-geneAnnot.csv", quote = FALSE, row.names = FALSE) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Save as a csv file.
```

### Decreases in maximum transcripts

#### Run enrichment test

```{r}
maxTransMaleNegDataBP <- new("topGOdata", ontology = "BP", allGenes = maxTransMaleNegGeneList,
                             annot = annFUN.gene2GO, gene2GO = geneID2GO) #Create biological process topGO object
maxTransMaleNegDataBP #Get summary of object
```

```{r}
test.stat <- new("classicCount", testStatistic = GOFisherTest, name = "Fisher test")
resultFisher.maxTransMaleNegDataBP <- getSigGroups(maxTransMaleNegDataBP, test.stat)
resultFisher.maxTransMaleNegDataBP
```

```{r}
pvalFis.maxTransMaleNegDataBP <- score(resultFisher.maxTransMaleNegDataBP) #Extract p-values
head(pvalFis.maxTransMaleNegDataBP)
hist(pvalFis.maxTransMaleNegDataBP, 50, xlab = "p-values") #Plot histogram of p-values
```

```{r}
allRes.maxTransMaleNegDataBP <- GenTable(maxTransMaleNegDataBP, classic = resultFisher.maxTransMaleNegDataBP, ranksOf = "classic", orderBy = "classic", topNodes = length(pvalFis.maxTransMaleNegDataBP))
#Create a statistical results table with statistical test results. Order by p-value (classic), and include all results (topNodes)
head(allRes.maxTransMaleNegDataBP)
```

```{r}
write.csv(allRes.maxTransMaleNegDataBP, "male-maxTransNeg-BP-FisherTestResults.csv", quote = FALSE, row.names = FALSE) #Save dataframe
```

#### Match enriched GOterms with general annotation information

```{r}
sigRes.maxTransMaleNegDataBP <- allRes.maxTransMaleNegDataBP[1:238,c(1, 6)] #Filter significantly enriched GOterms, only keep GOID and p-value
colnames(sigRes.maxTransMaleNegDataBP) <- c("GO", "p.value") #Change column names
head(sigRes.maxTransMaleNegDataBP)
```

```{r}
sigRes.maxTransMaleNegDataBP %>%
  left_join(., y = maxTransMaleAnnot, by = "GO") %>%
  filter(., difference < 0) %>%
  dplyr::select(., gene_name, difference, males_controls_max_transcript_counts, males_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  write.csv("male-maxTransNeg-BP-EnrichedGO-geneAnnot.csv", quote = FALSE, row.names = FALSE) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Save as a csv file.
```

# Changes in maximum transcripts expressed and methylation

## Format data

```{r}
maxTransMaleWide <- inner_join(male_tr, geneMethylationMaleWide, by = "gene_name") %>%
  inner_join(., geneExpressionMaleWide, by = "gene_name") #Join maximum transcript expression data with gene methylation and expression data
head(maxTransMaleWide)
```

## Methylation of genes with enriched BPGO

### Increases in maximum transcripts

```{r}
maxTransMalePosAnnot <- sigRes.maxTransMalePosDataBP %>%
  left_join(., y = maxTransMaleAnnot, by = "GO") %>%
  dplyr::select(., gene_name, difference, males_controls_max_transcript_counts, males_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  full_join(x = maxTransMaleWide, y = ., by = "gene_name") %>%
  dplyr::select(., gene_name, difference.x, change_meth, change_exp, GO, p.value, product) %>%
  dplyr::rename(., difference = "difference.x") %>%
  filter(., difference > 0) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Full join with the wide maximum transcript dataframe that has change in methylation and expression data. Select relevant columns and rename change column.
head(maxTransMalePosAnnot) #Confirm manipulation
```

```{r}
maleMaxTransPosEnrichedBPGO <- maxTransMalePosAnnot %>%
  filter(., is.na(GO) == FALSE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "", y = "Number of Genes", title = "A. Enriched BPGO") +
  theme_classic(base_size = 15)

maleMaxTransPosNotEnrichedBPGO <- maxTransMalePosAnnot %>%
  filter(., is.na(GO) == TRUE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "Change in Methylation (%)", y = "Number of Genes", title = "B. Not Enriched BPGO") +
  theme_classic(base_size = 15)
```

```{r}
maleMaxTransPosEnrichedBPGO / maleMaxTransPosNotEnrichedBPGO
ggsave("male-max-transcripts-positive-changeMeth-enrichedBPGO.pdf", height = 8.5, width = 11)
```

### Decreases in maximum transcripts

```{r}
maxTransMaleNegAnnot <- sigRes.maxTransMaleNegDataBP %>%
  left_join(., y = maxTransMaleAnnot, by = "GO") %>%
  dplyr::select(., gene_name, difference, males_controls_max_transcript_counts, males_exposed_max_transcript_counts, GO, p.value, product) %>%
  separate(., col = product, into = c("product", "transcriptVar"), sep = "%2C") %>%
  dplyr::select(., -transcriptVar) %>%
  unique(.) %>%
  full_join(x = maxTransMaleWide, y = ., by = "gene_name") %>%
  dplyr::select(., gene_name, difference.x, change_meth, change_exp, GO, p.value, product) %>%
  dplyr::rename(., difference = "difference.x") %>%
  filter(., difference < 0) #Take significant BPGO data and join with annotation information. Select columns of interest. Separate product column by "%2C" and remove transcriptVar column. Select unique rows. This removes all "transcript variant" information so each gene is listed once per significant GOterm. Full join with the wide maximum transcript dataframe that has change in methylation and expression data. Select relevant columns and rename change column.
head(maxTransMaleNegAnnot) #Confirm manipulation
```

```{r}
maleMaxTransNegEnrichedBPGO <- maxTransMaleNegAnnot %>%
  filter(., is.na(GO) == FALSE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "", y = "Number of Genes", title = "A. Enriched BPGO") +
  theme_classic(base_size = 15)

maleMaxTransNegNotEnrichedBPGO <- maxTransMaleNegAnnot %>%
  filter(., is.na(GO) == TRUE) %>%
  ggplot(aes(x = change_meth)) + 
  geom_histogram() +
  labs(x = "Change in Methylation (%)", y = "Number of Genes", title = "B. Not Enriched BPGO") +
  theme_classic(base_size = 15)
```

```{r}
maleMaxTransNegEnrichedBPGO / maleMaxTransNegNotEnrichedBPGO
ggsave("male-max-transcripts-negative-changeMeth-enrichedBPGO.pdf", height = 8.5, width = 11)
```

## Number of DML in genes with changes in the maximum transcripts expressed

```{r}
maxTransMaleWideDML <- left_join(maxTransMaleWide, DMLLocationsMale, by = "gene_name") %>%
  dplyr::select(., -c(chr, gene.start, gene.end)) %>%
  group_by(., gene_name) %>%
  mutate(., DMLcount = n()) %>%
  mutate(., DMLcount = case_when(is.na(meth.diff) == TRUE ~ 0,
                                 is.na(meth.diff) == FALSE ~ DMLcount)) #Join max trans data with DML annotations and remove superfluous columns. Group by gene name and count the number of DML per gene. When there is no DML in the gene (meth.diff == NA), change the DML count to 0.
head(maxTransMaleWideDML) #Confirm dataframe creation
```

```{r}
write.csv(maxTransMaleWideDML, "male-max-trans-DML-annotations.csv", quote = FALSE, row.names = FALSE) #Save dataframe
```

```{r}
maxTransMaleWideDML %>%
  group_by(., DMLcount, difference) %>%
  summarise(., count = n()) #Group data by DML count and change, then count the number of genes with or without predominant isoform shifts with various DML counts
```

DMLcount difference count
0	-7	1		
0	-5	2		
0	-4	3		
0	-3	13		
0	-2	87		
0	-1	863		
0	0	28314		
0	1	2134		
0	2	325		
0	3	72	
0	4	26		
0	5	13		
0	6	3		
0	7	3		
0	8	1		
0	13	1		
1	-3	1		
1	-2	7		
1	-1	57		
1	0	916	
1	1	95		
1	2	20		
1	3	5		
1	4	4		
1	6	1		
2	-2	2		
2	-1	16		
2	0	436		
2	1	58		
2	2	20	
2	3	2		
2	4	2		
2	7	2		
3	-3	3		
3	-2	6		
3	-1	18		
3	0	165		
3	1	21		
3	2	9		
3	3	3	
3	4	3		
4	-1	4		
4	0	144		
4	1	8		
4	2	4		
5	0	70		
5	3	10		
6	-2	6		
6	0	36		
7	0	28
7	3	7		
8	0	32		
9	0	27		
9	1	18		
11	0	11		
12	0	12		
13	0	13		
13	1	13		
14	-1	14		
15	-1	15
16	0	32		
17	-1	17		
23	0	23		
25	1	25	
