---
title: "78 asca meth-diff"
author: Steven Roberts
date: "`r format(Sys.time(), '%d %B, %Y')`" 
output: 
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  # github_document:
  #   toc: true
  #   toc_depth: 3
  #   number_sections: true
  #   html_preview: true
---

```{r setup, include=FALSE}
#library(kableExtra)
# library(DESeq2)
# library(pheatmap)
# library(RColorBrewer)
# library(data.table)
#library(DT)
# library(Biostrings)
#library(methylKit)
library(tidyverse)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,         # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  fig.width = 6,       # Set plot width in inches
  fig.height = 4,      # Set plot height in inches
  fig.align = "center" # Align plots to the center
)
```

Ariana created lists of genes were exon patterning was similar and DIFFERENT under OA conditions for both sexes.

The question becomes is there a clear difference in methylation patterns in these genes. 

This includes determining the preponderance in DMLs and

if gene body methylation differed.

## female where pattern differed.


```{r}
female_splice <- read.table("../output/77-asca-exon/females_splicing_treatments_gene_list.csv", col.names = c("name")) %>%
  mutate(desc = "f_diff")
```


female DMLs



```{r}
female_dml <- read.table("../output/16-DML-annotation/female_dml_geneID.txt") %>%
  mutate(desc = "fdml")
```



join

```{r}
fsp_dml <- female_splice %>%
  left_join(female_dml, by = c("name" = "V1"))
```

nothing


# GBM

```{r}
female_40_meth <- read_csv("../output/73-gene-methylation/female_40_meth.csv") %>%
  mutate(name = str_replace(name, "gene-", ""))
```


```{r}
f_splice_meth <- female_40_meth %>%
  left_join(female_splice, by = c("name" = "name")) %>%
  filter(desc == "f_diff")
```


# plot 


```{r}
ggplot(f_splice_meth, aes(x = average_control, y = average_exposed)) +
  geom_point() +  # This adds the points to the plot
  theme_minimal() 
```

## female no splice


```{r}
female_NOsplice <- read.table("../output/77-asca-exon/females_splicing_nodifference_gene_list.csv", col.names = c("name")) %>%
  mutate(desc = "f_same")
```


female DMLs



```{r}
female_dml <- read.table("../output/16-DML-annotation/female_dml_geneID.txt") %>%
  mutate(desc = "fdml")
```



join

```{r}
fNOsp_dml <- female_NOsplice %>%
  left_join(female_dml, by = c("name" = "V1"))
```

1 dml overlap


# GBM

```{r}
female_40_meth <- read_csv("../output/73-gene-methylation/female_40_meth.csv") %>%
  mutate(name = str_replace(name, "gene-", ""))
```


```{r}
f_NOsplice_meth <- female_40_meth %>%
  left_join(female_NOsplice, by = c("name" = "name")) %>%
  filter(desc == "f_same")
```


# plot 


```{r}
ggplot(f_NOsplice_meth, aes(x = average_control, y = average_exposed)) +
  geom_point() +  # This adds the points to the plot
  theme_minimal() 
```


# lets look at together on one plot

```{r}
ggplot() +
  geom_point(data = f_NOsplice_meth, aes(x = average_control, y = average_exposed), color = "blue") +
  geom_point(data = f_splice_meth, aes(x = average_control, y = average_exposed), color = "red") +
  theme_minimal() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_abline(intercept = -5, slope = 1, linetype = "solid", color = "grey") + # Line below
  geom_abline(intercept = 5, slope = 1, linetype = "solid", color = "grey") + # Line above
  labs(x = "control", y = "exposed", title = "female methylation difference")

```






--
Lets take make a histogram of methylation for those genes that change methylation versus those that do not.. Need to grab one version of the files Sam made and joing with recent (40-based) methylation data


This only has transcripts that change.

```{r}
female_tr <- read_csv("../output/34-transcript-counts/diffs.max.transcripts_per_gene.controls_females.vs.exposed_females.csv")
```



```{r}
fmcoe_max_predom_isos <- read_csv("../supplemental-files/fmcoe-max-predom-isos.csv")

```

lets join

```{r}
fmmax <- fmcoe_max_predom_isos %>%
  left_join(female_tr, by = "gene_name") %>%
  select(gene_name, control_females_max_transcript_counts.x, exposed_females_max_transcript_counts.x, difference)
```


lets join methylation data

```{r}
female_40_meth <- read_csv("../output/73-gene-methylation/female_40_meth.csv") %>%
  mutate(name = str_replace(name, "gene-", ""))
```
```{r}
structure(female_40_meth)
```


```{r}
col_name_1 <- names(fmmax)[1] # Get the name of the first column in fmmax
col_name_2 <- names(female_40_meth)[1] # Get the name of the first column in female_40_meth

# Perform the join
fmethmax <- left_join(female_40_meth,fmmax, by = setNames(col_name_1, col_name_2)) %>%
  mutate(mean_meth = (average_control + average_exposed)/2) %>%
  mutate(diff_meth = (average_exposed - average_control))
```



```{r}
# Add a new column to indicate whether 'difference' is NA or not
fmethmax$diff_status <- ifelse(is.na(fmethmax$difference), "NA", "Not NA")

# Plot
ggplot(fmethmax, aes(x = mean_meth, fill = diff_status)) +
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) + # Adjust binwidth as needed
  scale_fill_manual(values = c("NA" = "red", "Not NA" = "blue")) +
  labs(title = "Female gene methylation - change max transcript (not NA)",
       x = "Mean Methylation",
       fill = "Difference Status") +
  theme_minimal()
```
```{r}
ggplot(fmethmax, aes(x = mean_meth)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black") + # Adjust binwidth as needed
  facet_wrap(~diff_status, scales = "fixed") + 
  scale_y_log10() + # Apply log scale to y-axis) 
  labs(x = "Mean Meth", y = "Frequency") +
  theme_minimal() +
  ggtitle("Female gene methylation - change max transcript (not NA)")
```

# Plot



```{r}
ggplot(fmethmax, aes(x = mean_meth, fill = diff_status)) + 
  geom_histogram(binwidth = 1) + # Adjust binwidth as needed
  scale_fill_manual(values = c("NA" = "red", "Not NA" = "blue")) +
  scale_y_log10() + # Apply log scale to y-axis) 
  labs(x = "Mean Meth", y = "Frequency") +
  theme_minimal() +
  ggtitle("Female gene methylation - change max transcript (not NA)")
```

```{r}
ggplot(fmethmax, aes(x = difference, y = diff_meth)) +
  geom_point() +  # This adds the points to the plot
  theme_minimal() +  # Optional: Applies a minimalistic theme
  labs(x = "Max transcript Difference", y = "Meth Difference", title = "max transcript vs. methylation") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval
```

```{r}
ggplot(fmethmax, aes(x = difference, y = diff_meth)) +
  geom_jitter() +  # This replaces geom_point() and adds jitter to the points
  theme_minimal() +  # Applies a minimalistic theme
  labs(x = "Max transcript Difference", y = "Meth Difference", title = "max transcript vs. methylation") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval

```

```{r}
ggplot(fmethmax, aes(x = exposed_females_max_transcript_counts.x, y = average_exposed)) +
  geom_jitter() +  # This replaces geom_point() and adds jitter to the points
  theme_minimal() +  # Applies a minimalistic theme
  labs(x = "max transcript count", y = "gene methylation", title = "female exposed") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval

```




# male
```{r}
male_tr <- read_csv("../output/34-transcript-counts/diffs.max.transcripts_per_gene.controls_males.vs.exposed_males.csv")
```
```{r}
mmax <- fmcoe_max_predom_isos %>%
  left_join(male_tr, by = "gene_name") %>%
  select(gene_name, control_males_max_transcript_counts.x, exposed_males_max_transcript_counts.x, difference)
```


lets join methylation data

```{r}
male_40_meth <- read_csv("../output/73-gene-methylation/male_40_meth.csv") %>%
  mutate(name = str_replace(name, "gene-", ""))
```


```{r}
col_name_1 <- names(mmax)[1] # Get the name of the first column in fmmax
col_name_2 <- names(male_40_meth)[1] # Get the name of the first column in female_40_meth

# Perform the join
mmethmax <- left_join(male_40_meth,mmax, by = setNames(col_name_1, col_name_2)) %>%
  mutate(mean_meth = (average_control + average_exposed)/2) %>%
  mutate(diff_meth = (average_exposed - average_control))
```


```{r}
# Add a new column to indicate whether 'difference' is NA or not
mmethmax$diff_status <- ifelse(is.na(mmethmax$difference), "NA", "Not NA")

# Plot
ggplot(mmethmax, aes(x = mean_meth, fill = diff_status)) +
  geom_histogram(position = "identity", alpha = 0.5, binwidth = 1) + # Adjust binwidth as needed
  scale_fill_manual(values = c("NA" = "red", "Not NA" = "blue")) +
  labs(title = "Histogram of mean_meth",
       x = "Mean Methylation",
       fill = "Difference Status") +
  theme_minimal()
```


```{r}
ggplot(mmethmax, aes(x = mean_meth, fill = diff_status)) + 
  geom_histogram(binwidth = 1) + # Adjust binwidth as needed
  scale_fill_manual(values = c("NA" = "red", "Not NA" = "blue")) +
  scale_y_log10() + # Apply log scale to y-axis) 
  labs(x = "Mean Meth", y = "Frequency") +
  theme_minimal() +
  ggtitle("Histogram of Mean Meth by Difference Status")
```



```{r}
ggplot(mmethmax, aes(x = difference, y = diff_meth)) +
  geom_point() +  # This adds the points to the plot
  theme_minimal() +  # Optional: Applies a minimalistic theme
  labs(x = "Max transcript Difference", y = "Meth Difference", title = "max transcript vs. methylation") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval
```

```{r}
ggplot(mmethmax, aes(x = difference, y = diff_meth)) +
  geom_jitter() +  # This replaces geom_point() and adds jitter to the points
  theme_minimal() +  # Applies a minimalistic theme
  labs(x = "Max transcript Difference", y = "Meth Difference", title = "max transcript vs. methylation") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval

```



```{r}
ggplot(mmethmax, aes(x = exposed_males_max_transcript_counts.x, y = average_exposed)) +
  geom_jitter() +  # This replaces geom_point() and adds jitter to the points
  theme_minimal() +  # Applies a minimalistic theme
  labs(x = "max transcript count", y = "gene methylation", title = "male exposed") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval

```

```{r}
ggplot(mmethmax, aes(x = control_males_max_transcript_counts.x, y = average_control)) +
  geom_jitter() +  # This replaces geom_point() and adds jitter to the points
  theme_minimal() +  # Applies a minimalistic theme
  labs(x = "max transcript count", y = "gene methylation", title = "male control") +
  geom_smooth(method = "lm", se = FALSE, color = "blue")  # Adds a linear regression line without confidence interval

```