---
title: "Calculating FPKM coefficients of variation across sex/treatments in C.virginica gonad exposed to elevated pCO2 using Ballgown"
author: "Sam White"
date: "2/4/2022"
output: html_document
---
# Use [Ballgown](https://github.com/alyssafrazee/ballgown) to extract gene and transcript FPKM values and then calculate coefficients of variation across samples within treatments/sex of _C.virginica_ gonad tissue exposed to elevated pCO<sub>2</sub>.


REQUIRES the following R libraries:

- [`Ballgown`](https://github.com/alyssafrazee/ballgown) (Bioconductor)

- `tidyverse`

## Load `R` libraries

```{r}
library("ballgown")
library("tidyverse")
library("Rfast")
```

Notebooks detailing their creation:

- [FastQ trimming](https://robertslab.github.io/sams-notebook/2022/02/24/Trimming-Additional-20bp-from-C.virginica-Gonad-RNAseq-with-fastp-on-Mox.html)

- [Genome indexing, and exon/splice sites with HISAT2](https://robertslab.github.io/sams-notebook/2021/07/20/Genome-Annotations-Splice-Site-and-Exon-Extractions-for-C.virginica-GCF_002022765.2-Genome-Using-Hisat2-on-Mox.html)

- [Mapping and identificaion of isoforms with StingTie](https://robertslab.github.io/sams-notebook/2022/02/25/Transcript-Identification-and-Alignments-C.virginica-RNAseq-with-NCBI-Genome-GCF_002022765.2-Using-Hisat2-and-Stringtie-on-Mox.html)


## Load all transcript expression data

`TreatmentN` column explanation:

control males = 1
control females = 2
exposed males = 3
exposed females = 4
```{r}
# Expression data
whole_tx_table <-  read.csv("../data/whole_tx_table.csv")

# Rename gene_names listed as a "."
whole_tx_table <- whole_tx_table %>% mutate(gene_name = ifelse(gene_name == ".", t_name, gene_name))

head(whole_tx_table)
```

## Load all gene expression data
```{r load-gene-expression-data}
whole_gx_table <-  gexpr(bg)
head(whole_gx_table)
```


## Calculate coefficients of variation

`TreatmentN` column explanation:

control males = 1
control females = 2
exposed males = 3
exposed females = 4

```{r coefficients-of-variation}
# Vector of treatments
treatments <- c(1:4)

# Vector of comparisons
comparisons <- c("all", "f_v_m", "c_v_e")

# Create vector of ballgown gene names
## Moves the gene ids from rownames to a column in a data frame
ballgown_genes_df <- 
  as.data.frame(gexpr(bg)) %>%
  rownames_to_column(var = "gene_id")
## Extracts gene ids
genes <- ballgown_genes_df$gene_id

# Declare lists
transcripts_treatment_fpkm_list <- list()
genes_treatment_fpkm_list <- list()

# Remove data frame if it exists.
# Prevent problem of re-running and accidentally appending more data to existing data frame below.
rm(genes_control.vs.exposed_fpkm_CoV_df,
   genes_female.vs.male_fpkm_CoV_df,
   genes_treatment_fpkm_CoV_df,
   transcripts_control.vs.exposed_fpkm_CoV_df, 
   transcripts_female.vs.male_fpkm_CoV_df,
   transcripts_treatment_fpkm_CoV_df)

# Create list of FPKMs by treatment.
# Ordered in the list as listed above for the "TreatmentN" column explanation.
for (treatment in treatments) {
  
  # Create ballgown objects containing only data from individual treatments
  bg_transcripts <- ballgown::subset(bg, "TreatmentN == treatment", genomesubset = FALSE)
  
  # Create matrix containing gene FPKM from individual treatments
  bg_genes <- gexpr(ballgown::subset(bg, "TreatmentN == treatment", genomesubset = FALSE))
  
  
  # Extract transcript FPKMs and add to list
  transcripts_treatment_fpkm_list[[treatment]] <- texpr(bg_transcripts) 
  
  # Extract gene FPKMS and add to list
  genes_treatment_fpkm_list[[treatment]] <- bg_genes
}

for (comparison in comparisons) {
  
  # Create data frame with coefficients of variation for each Treatment group
  for (index in 1:length(transcripts_treatment_fpkm_list))  {
    if (comparison == "all") {
      # Set treatment column name
      treatment_col <- paste("Treatment", index, "_", "FPKM_Coeff_of_Var", sep = "")
    
      # Checks if data frame exists
      if (exists("transcripts_treatment_fpkm_CoV_df")) {
      
        # Add new column with coefficients of variation for each subsequent TreatmentN
        transcripts_treatment_fpkm_CoV_df$column <- rowcvs(transcripts_treatment_fpkm_list[[index]], unbiased = TRUE)
      } else
          # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
          transcripts_treatment_fpkm_CoV_df <- as.data.frame(rowcvs(transcripts_treatment_fpkm_list[[index]], unbiased = TRUE))
      
      # Checks if data frame exists
      if (exists("genes_treatment_fpkm_CoV_df")) {
      
        # Add new column with coefficients of variation for each subsequent TreatmentN
        genes_treatment_fpkm_CoV_df$column <- rowcvs(genes_treatment_fpkm_list[[index]], unbiased = TRUE)
      } else
          # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
          genes_treatment_fpkm_CoV_df <- as.data.frame(rowcvs(genes_treatment_fpkm_list[[index]], unbiased = TRUE))
    
      # Renames columns
      names(transcripts_treatment_fpkm_CoV_df)[index] <- treatment_col
      names(genes_treatment_fpkm_CoV_df)[index] <- treatment_col
    }
    
    # Treatments 2&4 (females)
    if (comparison == "f_v_m" & (index == "2" || index == "4")) {
      
      # If matrix doesn't exist, create it
      if (!exists("transcripts_females_matrix")) {
        transcripts_females_matrix <- transcripts_treatment_fpkm_list[[index]]
        
        # If matrix does exist, append column to matrix
      } else
          transcripts_females_matrix <- cbind(transcripts_treatment_fpkm_list[[index]])
      
      # If matrix doesn't exist, create it
      if (!exists("genes_females_matrix")) {
        genes_females_matrix <- genes_treatment_fpkm_list[[index]]
        
        # If matrix does exist, append column to matrix
      } else
          genes_females_matrix <- cbind(genes_treatment_fpkm_list[[index]])
      
      
      # Checks is data frame exists
      if (exists("transcripts_female.vs.male_fpkm_CoV_df")) {
      
        # Add new column with coefficients of variation for each subsequent TreatmentN
        transcripts_female.vs.male_fpkm_CoV_df$column <- rowcvs(transcripts_females_matrix, unbiased = TRUE)
      } else
          # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
          transcripts_female.vs.male_fpkm_CoV_df <- as.data.frame(rowcvs(transcripts_females_matrix, unbiased = TRUE))
      
      # Checks is data frame exists
      if (exists("genes_female.vs.male_fpkm_CoV_df")) {
      
        # Add new column with coefficients of variation for each subsequent TreatmentN
        genes_female.vs.male_fpkm_CoV_df$column <- rowcvs(genes_females_matrix, unbiased = TRUE)
      } else
          # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
          genes_female.vs.male_fpkm_CoV_df <- as.data.frame(rowcvs(genes_females_matrix, unbiased = TRUE))
      
    
    } # Treatments 1&3 (males)
      else if (comparison == "f_v_m" & (index == "1" || index == "3")) {
        
        # If matrix doesn't exist, create it
        if (!exists("transcripts_males_matrix")) {
          transcripts_males_matrix <- transcripts_treatment_fpkm_list[[index]]
          
          # If matrix does exist, append column to matrix
        } else
          transcripts_males_matrix <- cbind(transcripts_treatment_fpkm_list[[index]])
      
        # Checks if data frame exists
        if (exists("transcripts_female.vs.male_fpkm_CoV_df")) {
          
          # Add new column with coefficients of variation for each subsequent TreatmentN
          transcripts_female.vs.male_fpkm_CoV_df$column <- rowcvs(transcripts_males_matrix, unbiased = TRUE)
        } else
            # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
            transcripts_female.vs.male_fpkm_CoV_df <- as.data.frame(rowcvs(transcripts_males_matrix, unbiased = TRUE))

        # If matrix doesn't exist, create it
        if (!exists("genes_males_matrix")) {
          genes_males_matrix <- genes_treatment_fpkm_list[[index]]
          
          # If matrix does exist, append column to matrix
        } else
          genes_males_matrix <- cbind(genes_treatment_fpkm_list[[index]])
      
        # Checks if data frame exists
        if (exists("genes_female.vs.male_fpkm_CoV_df")) {
          
          # Add new column with coefficients of variation for each subsequent TreatmentN
          genes_female.vs.male_fpkm_CoV_df$column <- rowcvs(genes_males_matrix, unbiased = TRUE)
        } else
            # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
            genes_female.vs.male_fpkm_CoV_df <- as.data.frame(rowcvs(genes_males_matrix, unbiased = TRUE))

      
    } # Treatments 1&2 (control)
      else if (comparison == "c_v_e" & (index == "1" || index == "2")) {
        
        # If matrix doesn't exist, create it
        if (!exists("transcripts_controls_matrix")) {
          transcripts_controls_matrix <- transcripts_treatment_fpkm_list[[index]]
          
          # If matrix does exist, append column to matrix
        } else
            transcripts_controls_matrix <- cbind(transcripts_treatment_fpkm_list[[index]])
      
        # Checks if data frame exists
        if (exists("transcripts_control.vs.exposed_fpkm_CoV_df")) {
      
          # Add new column with coefficients of variation for each subsequent TreatmentN
          transcripts_control.vs.exposed_fpkm_CoV_df$column <- rowcvs(transcripts_controls_matrix, unbiased = TRUE)
        } else
            # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
            transcripts_control.vs.exposed_fpkm_CoV_df <- as.data.frame(rowcvs(transcripts_controls_matrix, unbiased = TRUE))
        
        # If matrix doesn't exist, create it
        if (!exists("genes_controls_matrix")) {
          genes_controls_matrix <- genes_treatment_fpkm_list[[index]]
          
          # If matrix does exist, append column to matrix
        } else
            genes_controls_matrix <- cbind(genes_treatment_fpkm_list[[index]])
      
        # Checks if data frame exists
        if (exists("genes_control.vs.exposed_fpkm_CoV_df")) {
      
          # Add new column with coefficients of variation for each subsequent TreatmentN
          genes_control.vs.exposed_fpkm_CoV_df$column <- rowcvs(genes_controls_matrix, unbiased = TRUE)
        } else
            # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
            genes_control.vs.exposed_fpkm_CoV_df <- as.data.frame(rowcvs(genes_controls_matrix, unbiased = TRUE))
      
    } # Treatments 3&4 (exposed)
      else if (comparison == "c_v_e" & (index == "3" || index == "4")) {
        
        # If matrix doesn't exist, create it
        if (!exists("transcripts_exposed_matrix")) {
          transcripts_exposed_matrix <- transcripts_treatment_fpkm_list[[index]]
          
          # If matrix does exist, append column to matrix
        } else
            transcripts_exposed_matrix <- cbind(transcripts_treatment_fpkm_list[[index]])
      
        # Checks if data frame exists
        if (exists("transcripts_control.vs.exposed_fpkm_CoV_df")) {
          
          # Add new column with coefficients of variation for each subsequent TreatmentN
          transcripts_control.vs.exposed_fpkm_CoV_df$column <- rowcvs(transcripts_exposed_matrix, unbiased = TRUE)
        } else
            # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
            transcripts_control.vs.exposed_fpkm_CoV_df <- as.data.frame(rowcvs(transcripts_exposed_matrix, unbiased = TRUE))
        
        # If matrix doesn't exist, create it
        if (!exists("genes_exposed_matrix")) {
          genes_exposed_matrix <- genes_treatment_fpkm_list[[index]]
          
          # If matrix does exist, append column to matrix
        } else
            genes_exposed_matrix <- cbind(genes_treatment_fpkm_list[[index]])
      
        # Checks if data frame exists
        if (exists("genes_control.vs.exposed_fpkm_CoV_df")) {
          
          # Add new column with coefficients of variation for each subsequent TreatmentN
          genes_control.vs.exposed_fpkm_CoV_df$column <- rowcvs(genes_exposed_matrix, unbiased = TRUE)
        } else
            # Creates initial data frame with coefficients of variation for FPKM in Treatment 1
            genes_control.vs.exposed_fpkm_CoV_df <- as.data.frame(rowcvs(genes_exposed_matrix, unbiased = TRUE))
    }
  }
}

# Rename columns
names(transcripts_female.vs.male_fpkm_CoV_df)[1] <- paste("males", "_", "FPKM_Coeff_of_Var", sep = "")
names(transcripts_female.vs.male_fpkm_CoV_df)[2] <- paste("females", "_", "FPKM_Coeff_of_Var", sep = "")
names(transcripts_control.vs.exposed_fpkm_CoV_df)[1] <- paste("controls", "_", "FPKM_Coeff_of_Var", sep = "")
names(transcripts_control.vs.exposed_fpkm_CoV_df)[2] <- paste("exposed", "_", "FPKM_Coeff_of_Var", sep = "")
names(genes_female.vs.male_fpkm_CoV_df)[1] <- paste("males", "_", "FPKM_Coeff_of_Var", sep = "")
names(genes_female.vs.male_fpkm_CoV_df)[2] <- paste("females", "_", "FPKM_Coeff_of_Var", sep = "")
names(genes_control.vs.exposed_fpkm_CoV_df)[1] <- paste("controls", "_", "FPKM_Coeff_of_Var", sep = "")
names(genes_control.vs.exposed_fpkm_CoV_df)[2] <- paste("exposed", "_", "FPKM_Coeff_of_Var", sep = "")

# Create gene ids column
genes_control.vs.exposed_fpkm_CoV_df <- cbind(gene_ids = genes, genes_control.vs.exposed_fpkm_CoV_df)
genes_female.vs.male_fpkm_CoV_df <- cbind(gene_ids = genes, genes_female.vs.male_fpkm_CoV_df)
genes_treatment_fpkm_CoV_df <- cbind(gene_ids = genes, genes_treatment_fpkm_CoV_df)

# Remove first row containing combo of unannotated transcripts
genes_control.vs.exposed_fpkm_CoV_df <- genes_control.vs.exposed_fpkm_CoV_df[-1,]
genes_female.vs.male_fpkm_CoV_df <- genes_female.vs.male_fpkm_CoV_df[-1,]
genes_treatment_fpkm_CoV_df <- genes_treatment_fpkm_CoV_df[-1,]


# Create transcript ID column
transcripts_control.vs.exposed_fpkm_CoV_df <- transcripts_control.vs.exposed_fpkm_CoV_df %>% 
  rownames_to_column(var = "t_id")

transcripts_female.vs.male_fpkm_CoV_df <- transcripts_female.vs.male_fpkm_CoV_df %>% 
  rownames_to_column(var = "t_id")

transcripts_treatment_fpkm_CoV_df <- transcripts_treatment_fpkm_CoV_df %>% 
  rownames_to_column(var = "t_id")

# Check genes data frames
head(genes_control.vs.exposed_fpkm_CoV_df)
head(genes_female.vs.male_fpkm_CoV_df)
head(genes_treatment_fpkm_CoV_df)

# Check transcripts data frames
head(transcripts_control.vs.exposed_fpkm_CoV_df)
head(transcripts_female.vs.male_fpkm_CoV_df)
head(transcripts_treatment_fpkm_CoV_df)



# Write genes data frames to CSVs
write.csv(genes_control.vs.exposed_fpkm_CoV_df,
          file = "../analyses/genes_control.vs.exposed_fpkm_CoV_df.csv",
          row.names = FALSE,
          quote = FALSE)

write.csv(genes_female.vs.male_fpkm_CoV_df,
          file = "../analyses/genes_female.vs.male_fpkm_CoV_df.csv",
          row.names = FALSE,
          quote = FALSE)

write.csv(genes_treatment_fpkm_CoV_df,
          file = "../analyses/genes_treatment_fpkm_CoV_df.csv",
          row.names = FALSE,
          quote = FALSE)

# Write transcripts data frames to CSVs
write.csv(transcripts_control.vs.exposed_fpkm_CoV_df,
          file = "../analyses/transcripts_control.vs.exposed_fpkm_CoV_df.csv",
          row.names = FALSE,
          quote = FALSE)

write.csv(transcripts_female.vs.male_fpkm_CoV_df,
          file = "../analyses/transcripts_female.vs.male_fpkm_CoV_df.csv",
          row.names = FALSE,
          quote = FALSE)

write.csv(transcripts_treatment_fpkm_CoV_df,
          file = "../analyses/transcripts_treatment_fpkm_CoV_df.csv",
          row.names = FALSE,
          quote = FALSE)


```
