---
title: "General Methylation Landscape"
author: "Yaamini Venkataraman"
output: github_document
---

In this script, I'll create summary tables for genomic location information, run statistcal tests, and visualize the distribution of DML in the genome.

# Prepare R Markdown file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/yaaminivenkataraman/Documents/ceabigr/output/methylation-landscape/") #Set root directory
```

# Install packages

```{r}
#Packages for visualization

#install.packages("dichromat")
#install.packages("broom")

require(dichromat)
require(broom)
```

# Obtain session information

```{r}
sessionInfo()
```

# Establish color scheme

```{r}
plotColors <- rev(RColorBrewer::brewer.pal(5, "Blues")) #Create a color palette for the barplots. Use 5 shades from RColorBrewer. Reverse the order so the darkest shade is used first.
```

```{r}
plotColors2 <- rev(RColorBrewer::brewer.pal(8, "Blues")) #Create a color palette for the barplots. Use 8 shades from RColorBrewer. Reverse the order so the darkest shade is used first.
```

# Import file counts

```{r}
exonUTROverlaps <- read.table("exonUTR-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femExonUTROverlaps <- exonUTROverlaps[c(80,82,81),] #Female data, reorder rows
maleExonUTROverlaps <- exonUTROverlaps[c(84,86,85),] #Female data, reorder rows
head(femExonUTROverlaps) #Confirm dataframe creation
head(maleExonUTROverlaps) #Confirm dataframe creation
```

```{r}
CDSOverlaps <- read.table("CDS-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femCDSOverlaps <- CDSOverlaps[c(80,82,81),] #Female data, reorder rows
maleCDSOverlaps <- CDSOverlaps[c(84,86,85),] #Female data, reorder rows
head(femCDSOverlaps) #Confirm dataframe creation
head(maleCDSOverlaps) #Confirm dataframe creation
```

```{r}
intronOverlaps <- read.table("intron-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femIntronOverlaps <- intronOverlaps[c(80,82,81),] #Female data, reorder rows
maleIntronOverlaps <- intronOverlaps[c(84,86,85),] #Female data, reorder rows
head(femIntronOverlaps) #Confirm dataframe creation
head(maleIntronOverlaps) #Confirm dataframe creation
```

```{r}
upstreamOverlaps <- read.table("upstreamFlanks-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femUpstreamOverlaps <- upstreamOverlaps[c(80,82,81),] #Female data, reorder rows
maleUpstreamOverlaps <- upstreamOverlaps[c(84,86,85),] #Female data, reorder rows
head(femUpstreamOverlaps) #Confirm dataframe creation
head(maleUpstreamOverlaps) #Confirm dataframe creation
```

```{r}
downstreamOverlaps <- read.table("downstreamFlanks-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femDownstreamOverlaps <- downstreamOverlaps[c(80,82,81),] #Female data, reorder rows
maleDownstreamOverlaps <- downstreamOverlaps[c(84,86,85),] #Female data, reorder rows
head(femDownstreamOverlaps) #Confirm dataframe creation
head(maleDownstreamOverlaps) #Confirm dataframe creation
```

```{r}
lncRNAOverlaps <- read.table("lncRNA-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femlncRNAOverlaps <- lncRNAOverlaps[c(80,82,81),] #Female data, reorder rows
malelncRNAOverlaps <- lncRNAOverlaps[c(84,86,85),] #Female data, reorder rows
head(femlncRNAOverlaps) #Confirm dataframe creation
head(malelncRNAOverlaps) #Confirm dataframe creation
```

```{r}
TEOverlaps <- read.table("TE-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femTEOverlaps <- TEOverlaps[c(80,82,81),] #Female data, reorder rows
maleTEOverlaps <- TEOverlaps[c(84,86,85),] #Female data, reorder rows
head(femTEOverlaps) #Confirm dataframe creation
head(maleTEOverlaps) #Confirm dataframe creation
```

```{r}
intergenicOverlaps <- read.table("intergenic-counts.txt", header = FALSE, col.names = c("counts", "filename")) #Import file with all file overlaps
femIntergenicOverlaps <- intergenicOverlaps[c(80,82,81),] #Female data, reorder rows
maleIntergenicOverlaps <- intergenicOverlaps[c(84,86,85),] #Female data, reorder rows
head(femIntergenicOverlaps) #Confirm dataframe creation
head(maleIntergenicOverlaps) #Confirm dataframe creation
```

# Females

## Genomic location of methylated CpGs

My goal is to 1) understand statistical differences between distribution of CG motifs and methylated CpGs in various genomic features and 2) visualize the genomic locations of these loci.

### Create summary table

```{r}
femCpGFeatureOverlaps <- cbind(femExonUTROverlaps,
                               femCDSOverlaps,
                               femIntronOverlaps,
                               femUpstreamOverlaps,
                               femDownstreamOverlaps,
                               femlncRNAOverlaps,
                               femTEOverlaps,
                               femIntergenicOverlaps) #Combine information from all genome features
rownames(femCpGFeatureOverlaps) <- c("Meth", "modMeth", "lowMeth") #Assign row names
ncol(femCpGFeatureOverlaps) #Count columns
femCpGFeatureOverlaps <- femCpGFeatureOverlaps[,seq(1, 16, 2)] #Keep odd-numbered columns (counts)
colnames(femCpGFeatureOverlaps) <- c("exonUTR", "CDS", "intron", "upstream", "downstream", "lncRNA", "TE", "intergenic") #Add column names
head(femCpGFeatureOverlaps) #Confirm formatting
```

```{r}
femCpGFeatureOverlaps <- as.data.frame(t(femCpGFeatureOverlaps)) #Transpose so column names are row names
femCpGFeatureOverlaps$allCpGs <- c(606308, 1728303, 5497597, 694265, 616684, 281715, 611471, 5417334) #Add CpG motif overlap information
femCpGFeatureOverlaps <- femCpGFeatureOverlaps[,c(4, 1:3)] #Reorganize columns
head(femCpGFeatureOverlaps) #Confirm formatting
```

```{r}
write.csv(femCpGFeatureOverlaps, "fem-CpG-feature-overlaps.csv", quote = FALSE, row.names = TRUE) #Save data
```

### Contingency tests

#### Format data

```{r}
femCpGLocationStatTest <- data.frame(t(femCpGFeatureOverlaps)) #Transpose for statistical testing
femCpGLocationStatTest <- femCpGLocationStatTest[1:2,] #keep only allCpGs and Meth data
head(femCpGLocationStatTest) #Confirm formatting
```

#### All vs. Meth

```{r}
femCpGLocationAllMeth <- data.frame() #Create empty dataframe to store chi-squared results
for(i in 1:ncol(femCpGLocationStatTest)) { #For each genome feature
  AllFeature <- femCpGLocationStatTest[1,i] #Variable for # genome feature overlaps for genomic CpGs with data
  MethFeature <- femCpGLocationStatTest[2,i] #Variable for # genome feature overlaps for methylated CpGs
  AllNotFeature <- sum(femCpGLocationStatTest[1,-i]) #Variable for # other CpG types for genomic CpGs with data
  MethNotFeature <- sum(femCpGLocationStatTest[2,-i]) #Variable for # other CpG types for methylated CpGs
  ct <- matrix(c(AllFeature, MethFeature, AllNotFeature, MethNotFeature), ncol = 2) #Create contingency table
  colnames(ct) <- c(as.character(colnames(femCpGLocationStatTest[i])), paste0("Not", colnames(femCpGLocationStatTest[i]))) #Assign column names: type, not type
  rownames(ct) <- c(as.character(row.names(femCpGLocationStatTest)[c(1,2)])) #Assign row names: genomic CpGs with data, methylated CpGs
  print(ct) #Confirm table is correct
  ctResults <- data.frame(broom::tidy(chisq.test(ct))) #Create dataframe storing chi-sq stats results. Use broom::tidy to extract results from test output
  ctResults$GenomeFeature <- as.character(colnames(femCpGLocationStatTest)[i]) #Add CpG type to results
  femCpGLocationAllMeth <- rbind(femCpGLocationAllMeth, ctResults) #Add test statistics to master table
}
```

```{r}
head(femCpGLocationAllMeth)
```

```{r}
femCpGLocationAllMeth$p.adj <- p.adjust(femCpGLocationAllMeth$p.value, method = "fdr") #Correct p-value using FDR
range(femCpGLocationAllMeth$p.adj) #Look at range of p-values
head(femCpGLocationAllMeth) #Confirm changes
```

```{r}
write.csv(femCpGLocationAllMeth, "fem-CpG-location-statResults.txt", quote = FALSE, row.names = TRUE) #Save statistical output
```

### Create barplot

```{r}
head(femCpGFeatureOverlaps)
```

```{r}
femCpGFeatureOverlapsPercents <- femCpGFeatureOverlaps #Duplicate dataframe
for (i in 1:length(femCpGFeatureOverlaps)) {
  femCpGFeatureOverlapsPercents[,i] <- (femCpGFeatureOverlapsPercents[,i] / (sum(femCpGFeatureOverlapsPercents[,i]))) * 100
} #Divide every entry by sum of the column and multiply by 100 to get percentages. Do not include gene information
head(femCpGFeatureOverlapsPercents) #Check calculations
```

```{bash}
mkdir figures #Make figure directory
```

```{r}
#pdf("figures/fem-CpG-feature-overlaps.pdf", width = 11, height = 8.5)

par(mar = c(1,5,0,1), oma = c(3, 1, 1, 11)) #Change figure boundaries

barplot(t(t(femCpGFeatureOverlapsPercents[,c(1:4)])), 
        col= plotColors2, 
        axes = FALSE, 
        names.arg = c("5x CpGs", "High", "Moderate", "Low"), cex.names = 1.5,
        ylim = c(0, 110)) #Create base plot. Everything should be white. Do not plot axes. Include sequencing type as labels and set size. Set y-axis specs. 

axis(side = 2, at = seq(0, 100, by = 10), las = 2, col = "grey80", cex.axis = 1.2) #Add y-axis
mtext(side = 2, "% CpGs", line = 3, cex = 1.5) #Add y-axis label

par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE) #Create new plot
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n") #Add new plot on top of current plot
legend(x = 0.57, y = 0.87, 
       xpd = TRUE,
       legend = c("Exon UTR", 
                  "CDS", 
                  "Introns", 
                  "Upstream Flank", 
                  "Downstream Flank", 
                  "lncRNA", 
                  "TE", 
                  "Intergenic"),
       pch = 22, 
        col = "black", 
        pt.bg = plotColors2,
       bty = "n",
       cex = 1.5, 
       x.intersp = 0.7, xjust = 0) #Place a legend in the top right of the figure with no box
text("Genome Feature", x = 0.785, y = 0.879, cex = 1.5) #Add legend title that is aligned with legend

#dev.off()
```

# Males

## Genomic location of methylated CpGs

### Create summary table

```{r}
maleCpGFeatureOverlaps <- cbind(maleExonUTROverlaps,
                               maleCDSOverlaps,
                               maleIntronOverlaps,
                               maleUpstreamOverlaps,
                               maleDownstreamOverlaps,
                               malelncRNAOverlaps,
                               maleTEOverlaps,
                               maleIntergenicOverlaps) #Combine information from all genome features
rownames(maleCpGFeatureOverlaps) <- c("Meth", "modMeth", "lowMeth") #Assign row names
ncol(maleCpGFeatureOverlaps) #Count columns
maleCpGFeatureOverlaps <- maleCpGFeatureOverlaps[,seq(1, 16, 2)] #Keep odd-numbered columns (counts)
colnames(maleCpGFeatureOverlaps) <- c("exonUTR", "CDS", "intron", "upstream", "downstream", "lncRNA", "TE", "intergenic") #Add column names
head(maleCpGFeatureOverlaps) #Confirm formatting
```

```{r}
maleCpGFeatureOverlaps <- as.data.frame(t(maleCpGFeatureOverlaps)) #Transpose so column names are row names
maleCpGFeatureOverlaps$allCpGs <- c(606308, 1728303, 5497597, 694265, 616684, 281715, 611471, 5417334) #Add CpG motif overlap information
maleCpGFeatureOverlaps <- maleCpGFeatureOverlaps[,c(4, 1:3)] #Reorganize columns
head(maleCpGFeatureOverlaps) #Confirm formatting
```

```{r}
write.csv(maleCpGFeatureOverlaps, "male-CpG-feature-overlaps.csv", quote = FALSE, row.names = TRUE) #Save data
```

### Contingency tests

#### Format data

```{r}
maleCpGLocationStatTest <- data.frame(t(maleCpGFeatureOverlaps)) #Transpose for statistical testing
maleCpGLocationStatTest <- maleCpGLocationStatTest[1:2,] #keep only allCpGs and Meth data
head(maleCpGLocationStatTest) #Confirm formatting
```

#### All vs. Meth

```{r}
maleCpGLocationAllMeth <- data.frame() #Create empty dataframe to store chi-squared results
for(i in 1:ncol(maleCpGLocationStatTest)) { #For each genome feature
  AllFeature <- maleCpGLocationStatTest[1,i] #Variable for # genome feature overlaps for genomic CpGs with data
  MethFeature <- maleCpGLocationStatTest[2,i] #Variable for # genome feature overlaps for methylated CpGs
  AllNotFeature <- sum(maleCpGLocationStatTest[1,-i]) #Variable for # other CpG types for genomic CpGs with data
  MethNotFeature <- sum(maleCpGLocationStatTest[2,-i]) #Variable for # other CpG types for methylated CpGs
  ct <- matrix(c(AllFeature, MethFeature, AllNotFeature, MethNotFeature), ncol = 2) #Create contingency table
  colnames(ct) <- c(as.character(colnames(maleCpGLocationStatTest[i])), paste0("Not", colnames(maleCpGLocationStatTest[i]))) #Assign column names: type, not type
  rownames(ct) <- c(as.character(row.names(maleCpGLocationStatTest)[c(1,2)])) #Assign row names: genomic CpGs with data, methylated CpGs
  print(ct) #Confirm table is correct
  ctResults <- data.frame(broom::tidy(chisq.test(ct))) #Create dataframe storing chi-sq stats results. Use broom::tidy to extract results from test output
  ctResults$GenomeFeature <- as.character(colnames(maleCpGLocationStatTest)[i]) #Add CpG type to results
  maleCpGLocationAllMeth <- rbind(maleCpGLocationAllMeth, ctResults) #Add test statistics to master table
}
```

```{r}
head(maleCpGLocationAllMeth)
```

```{r}
maleCpGLocationAllMeth$p.adj <- p.adjust(maleCpGLocationAllMeth$p.value, method = "fdr") #Correct p-value using FDR
range(maleCpGLocationAllMeth$p.adj) #Look at range of p-values
head(maleCpGLocationAllMeth) #Confirm changes
```

```{r}
write.csv(maleCpGLocationAllMeth, "male-CpG-location-statResults.txt", quote = FALSE, row.names = TRUE) #Save statistical output
```

### Create barplot

```{r}
head(maleCpGFeatureOverlaps)
```

```{r}
maleCpGFeatureOverlapsPercents <- maleCpGFeatureOverlaps #Duplicate dataframe
for (i in 1:length(maleCpGFeatureOverlaps)) {
  maleCpGFeatureOverlapsPercents[,i] <- (maleCpGFeatureOverlapsPercents[,i] / (sum(maleCpGFeatureOverlapsPercents[,i]))) * 100
} #Divide every entry by sum of the column and multiply by 100 to get percentages. Do not include gene information
head(maleCpGFeatureOverlapsPercents) #Check calculations
```

```{r}
#pdf("figures/male-CpG-feature-overlaps.pdf", width = 11, height = 8.5)

par(mar = c(1,5,0,1), oma = c(3, 1, 1, 11)) #Change figure boundaries

barplot(t(t(maleCpGFeatureOverlapsPercents[,c(1:4)])), 
        col= plotColors2, 
        axes = FALSE, 
        names.arg = c("5x CpGs", "High", "Moderate", "Low"), cex.names = 1.5,
        ylim = c(0, 110)) #Create base plot. Everything should be white. Do not plot axes. Include sequencing type as labels and set size. Set y-axis specs. 

axis(side = 2, at = seq(0, 100, by = 10), las = 2, col = "grey80", cex.axis = 1.2) #Add y-axis
mtext(side = 2, "% CpGs", line = 3, cex = 1.5) #Add y-axis label

par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE) #Create new plot
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n") #Add new plot on top of current plot
legend(x = 0.57, y = 0.87, 
       xpd = TRUE,
       legend = c("Exon UTR", 
                  "CDS", 
                  "Introns", 
                  "Upstream Flank", 
                  "Downstream Flank", 
                  "lncRNA", 
                  "TE", 
                  "Intergenic"),
       pch = 22, 
        col = "black", 
        pt.bg = plotColors2,
       bty = "n",
       cex = 1.5, 
       x.intersp = 0.7, xjust = 0) #Place a legend in the top right of the figure with no box
text("Genome Feature", x = 0.785, y = 0.879, cex = 1.5) #Add legend title that is aligned with legend

#dev.off()
```
