---
title: "exon_coverage"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Coverage of exons in gene expression data.  

Download sorted bam files.  
```{bash}
wget -r --no-check-certificate --quiet --no-directories --no-parent -P ../../data/SortedBAM -A *sorted.bam https://gannet.fish.washington.edu/Atumefaciens/20210726_cvir_stringtie_GCF_002022765.2_isoforms/
```

Run bedtools to generate coverage information for each sample.    
```{bash}
cd ~/Projects/ceabigr/data/SortedBAM

FILES=$(ls *.sorted.bam)

for file in ${FILES}
do
    NAME=$(echo ${file} | awk -F "." '{print $1}')
    echo ${NAME}

    /home/shared/bedtools2/bin/bedtools coverage \
    -mean \
    -sorted \
    -a ~/Projects/ceabigr/genome-features/C_virginica-3.0_Gnomon_exon_sorted_yrv.bed \
    -b ~/Projects/ceabigr/data/SortedBAM/${NAME}.sorted.bam \
    > ~/Projects/ceabigr/data/exon_coverage/${NAME}_exon_cov_m.out
done
```

# Coverage of TE in gene expression data.  

Run .bam files against C_virginica-3.0_TE-all.gff -> turn in .bed files 

```{bash}
cd ~/Projects/ceabigr/genome-features/

awk 'BEGIN{FS=OFS="\t"} {print $1,$4,$5}' C_virginica-3.0_TE-all.gff > C_virginica-3.0_TE-all.bed

```

Run this against the TE files to get out files for TE 

Run bedtools to generate coverage information for each sample.    
```{bash}
cd ~/Projects/ceabigr/data/SortedBAM

FILES=$(ls *.sorted.bam)

for file in ${FILES}
do
    NAME=$(echo ${file} | awk -F "." '{print $1}')
    echo ${NAME}

    /home/shared/bedtools2/bin/bedtools coverage \
    -mean \
    -sorted \
    -a ~/Projects/ceabigr/genome-features/C_virginica-3.0_TE-all.bed \
    -b ~/Projects/ceabigr/data/SortedBAM/${NAME}.sorted.bam \
    > ~/Projects/ceabigr/data/TE_coverage/${NAME}_TE_cov_m.out
done
```

# Trial correlation of gene expression coverage and methylation of TE's in one sample 

Run in sample S3F between TE coverage and TE methylation

Join together a test file with TE methylation adding the coverage column from our TE coverage file 

TE coverage: ~Projects/ceabigr/data/TE_coverage/S3F_TE_cov_m.out
TE methylation: ~Projects/ceabigr/output/3F_R1_val_1_bismark_bt2_pe_te-allcov_m.out
```{r}
library(tidyverse)
TE_genes<-read.csv("~/Projects/ceabigr/data/TE_coverage/S6M_TE_cov_m.out", sep="\t", header=FALSE)
TE_meth<-read.csv("~/Projects/ceabigr/output/6M_R1_val_1_bismark_bt2_pe_te-allcov_m.out", sep="\t", header=FALSE)

colnames(TE_genes)<-c("chromosome", "start", "end", "gene_cov")
colnames(TE_meth)<-c("chromosome", "predmodel", "calc", "start", "end", "X", "Y", "Z", "info", "meth_cov")

test<-left_join(TE_meth, TE_genes)
```

Run correlation analysis 

```{r}
test<-test%>%
    filter(meth_cov<400)

plot(test$gene_cov~test$meth_cov)
abline(lm(test$gene_cov~test$meth_cov), data=test, col="red")

cor.test(test$gene_cov,test$meth_cov)
```

# TE coverage joining files

Genes: 
```{r}
library(purrr)
library(fs)

# binding all gene coverage data files together
data_genes_TE <- list.files(path = '~/Projects/ceabigr/data/TE_coverage', pattern = ".out", full.names = TRUE)%>%
    set_names(.) %>%
    map_dfr(read.table, .id = "file.ID", header=FALSE, sep="\t")

data_genes_TE$sample_id<-sub(".*coverage/S", "", data_genes_TE$file.ID) #grab after files/
data_genes_TE$sample_id<-sub("_TE.*", "", data_genes_TE$sample_id) #grab before underscore

data_genes_TE<-data_genes_TE%>%
    select(!file.ID)

data_genes_TE<-data_genes_TE%>%
    mutate(sex = case_when(
        endsWith(sample_id, "M") ~ "Male",
        endsWith(sample_id, "F") ~ "Female"
        ))

colnames(data_genes_TE)<-c("chromosome", "start", "end", "genes", "sample_id", "sex")

#write this to a file
write.csv(data_genes_TE, "~/Projects/ceabigr/output/gene_coverage_TE.csv")
```

Methylation:
```{r}
# binding all gene coverage data files together
data_meth_TE <- list.files(path = '/home/shared/8TB_HDD_02/javirodr/ceabigr/data/TE_analysis/meth_TE_files', pattern = ".cov", full.names = TRUE)%>%
    set_names(.) %>%
    map_dfr(read.table, .id = "file.ID", header=FALSE, sep="\t")

#summarize mean of V4 (methylation level) and sum of V5 (methylation density) by unique combinations of V6, V9, V10 
data_meth_TE$code<-paste(data_meth_TE$V6, data_meth_TE$V9, data_meth_TE$V10)

data_meth_TE_sum<-data_meth_TE%>%
    group_by(code)%>%
    summarise(level=mean(V4, na.rm=TRUE), 
              coverage=sum(V5, na.rm=TRUE))

data_meth_TE_sum$chromosome<-data_meth_TE$V6[match(data_meth_TE_sum$code, data_meth_TE$code)]
data_meth_TE_sum$start<-data_meth_TE$V9[match(data_meth_TE_sum$code, data_meth_TE$code)]
data_meth_TE_sum$end<-data_meth_TE$V10[match(data_meth_TE_sum$code, data_meth_TE$code)]
data_meth_TE_sum$file.ID<-data_meth_TE$file.ID[match(data_meth_TE_sum$code, data_meth_TE$code)]

data_meth_TE_sum$sample_id<-sub(".*files/", "", data_meth_TE_sum$file.ID) #grab after files/
data_meth_TE_sum$sample_id<-sub("_R.*", "", data_meth_TE_sum$sample_id) #grab before underscore
#want to grab after "files/" and before "_R"

data_meth_TE_sum<-data_meth_TE_sum%>%
    select(!file.ID) %>%
    select(!code)#may need to remove code

data_meth_TE_sum<-data_meth_TE_sum%>%
    mutate(sex = case_when(
        endsWith(sample_id, "M") ~ "Male",
        endsWith(sample_id, "F") ~ "Female"
        ))

#write this to a file
write.csv(data_meth_TE_sum, "~/Projects/ceabigr/output/methylation_level_TE.csv")
```

Join files together
```{r}
TE_file<-left_join(data_genes_TE, data_meth_TE_sum)
head(TE_file)
```

Plot correlation.  

```{r}
library(ggplot2)

TE_plot<-TE_file%>%
    ggplot(., aes(x=level, y=genes, color=sex))+
    #geom_smooth(method="loess", se=TRUE, fullrange=TRUE, level=0.95, color="black") +
    geom_point(pch = 21, size=4) + 
    xlab("Methylation level") + 
    ylab("Gene expression")+
    ylim(0,300)+
    xlim(0,150)+
    theme_classic() + 
    theme(
      legend.position="right",
      axis.title=element_text(face="bold", size=14),
      axis.text=element_text(size=12, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=12)
      ); TE_plot

```

NEXT NEED TO PLOT METHYLATION COVERAGE AGAINST METHYLATION LEVEL

# Join files for exon level 

```{r}
library(purrr)
library(fs)

# binding all gene coverage data files together
data_genes_exon <- list.files(path = '~/Projects/ceabigr/data/exon_coverage', pattern = ".out", full.names = TRUE)%>%
    set_names(.) %>%
    map_dfr(read.table, .id = "file.ID", header=FALSE, sep="\t")

data_genes_exon$sample_id<-substr(data_genes_exon$file.ID,69,71)

data_genes_exon<-data_genes_exon%>%
    select(!file.ID)

data_genes_exon<-data_genes_exon%>%
    mutate(sex = case_when(
        endsWith(sample_id, "M") ~ "Male",
        endsWith(sample_id, "F") ~ "Female"
        ))
    
colnames(data_genes_exon)<-c("chromosome", "start", "end", "gene_cov", "sample_id", "sex")

#write this to a file
write.csv(data_genes_exon, "~/Projects/ceabigr/output/gene_coverage_exon.csv")
```

Genes: 
```{r}
# binding all gene coverage data files together
data_genes_exon <- list.files(path = '~/Projects/ceabigr/data/exon_coverage', pattern = ".out", full.names = TRUE)%>%
    set_names(.) %>%
    map_dfr(read.table, .id = "file.ID", header=FALSE, sep="\t")

data_genes_exon$sample_id<-sub(".*coverage/S", "", data_genes_exon$file.ID) #grab after files/
data_genes_exon$sample_id<-sub("_exon.*", "", data_genes_exon$sample_id) #grab before underscore

data_genes_exon<-data_genes_exon%>%
    select(!file.ID)

data_genes_exon<-data_genes_exon%>%
    mutate(sex = case_when(
        endsWith(sample_id, "M") ~ "Male",
        endsWith(sample_id, "F") ~ "Female"
        ))

colnames(data_genes_exon)<-c("chromosome", "start", "end", "genes", "sample_id", "sex")

#write this to a file
write.csv(data_genes_exon, "~/Projects/ceabigr/output/gene_coverage_exon.csv")
```

Methylation:
```{r}
# binding all methylation data files together
data_meth_exon <- list.files(path = '/home/shared/8TB_HDD_02/javirodr/ceabigr/data/TE_analysis/meth_exon_files', pattern = ".cov", full.names = TRUE)%>%
    set_names(.) %>%
    map_dfr(read.table, .id = "file.ID", header=FALSE, sep="\t")

#paste together scaffold exon coordinates into a "code"
data_meth_exon$code<-paste(data_meth_exon$V6, data_meth_exon$V7, data_meth_exon$V8)

#summarize mean of V4 (methylation level) and sum of V5 (methylation density) by unique combinations of V6, V7, V8 
data_meth_exon_sum<-data_meth_exon%>%
    group_by(code)%>%
    summarise(level=mean(V4, na.rm=TRUE), 
              coverage=sum(V5, na.rm=TRUE))

data_meth_exon_sum$chromosome<-data_meth_exon$V6[match(data_meth_exon_sum$code, data_meth_exon$code)]
data_meth_exon_sum$start<-data_meth_exon$V7[match(data_meth_exon_sum$code, data_meth_exon$code)]
data_meth_exon_sum$end<-data_meth_exon$V8[match(data_meth_exon_sum$code, data_meth_exon$code)]
data_meth_exon_sum$file.ID<-data_meth_exon$file.ID[match(data_meth_exon_sum$code, data_meth_exon$code)]

data_meth_exon_sum$sample_id<-sub(".*files/", "", data_meth_exon_sum$file.ID) #grab after files/
data_meth_exon_sum$sample_id<-sub("_R.*", "", data_meth_exon_sum$sample_id) #grab before underscore
#want to grab after "files/" and before "_R"

data_meth_exon_sum<-data_meth_exon_sum%>%
    select(!file.ID) %>%
    select(!code)#may need to remove code

data_meth_exon_sum<-data_meth_exon_sum%>%
    mutate(sex = case_when(
        endsWith(sample_id, "M") ~ "Male",
        endsWith(sample_id, "F") ~ "Female"
        ))

#remove na's
data_meth_exon_sum<-data_meth_exon_sum%>%
  filter(!is.na(chromosome))
```

```{r}
#write this to a file
write.csv(data_meth_exon_sum, "~/Projects/ceabigr/output/methylation_level_exon.csv")
```

Load files if starting from this point.  

```{r}
library(tidyverse)
data_meth_exon_sum<-read.csv("~/Projects/ceabigr/output/methylation_level_exon.csv", sep=",")

data_genes_exon<-read.csv("~/Projects/ceabigr/output/gene_coverage_exon.csv", sep=",")

head(data_meth_exon_sum)
data_meth_exon_sum<-data_meth_exon_sum%>%
  select(!X)

head(data_genes_exon)
data_genes_exon<-data_genes_exon%>%
  select(!X)
```

Load metadata file.  
```{r}
metadata<-read.csv(file="~/Projects/ceabigr/data/adult-meta.csv", sep=",", header=TRUE)

metadata<-metadata%>%
  select(Sample.ID, Treatment)%>%
  rename(sample_id=Sample.ID)
```

Join files together
```{r}
exon_file<-left_join(data_meth_exon_sum, data_genes_exon)

exon_file<-exon_file%>%
  unique(.)

exon_file<-left_join(exon_file, metadata)

head(exon_file)
```

Plot correlation.  

```{r}
exon_plot<-exon_file%>%
    ggplot(., aes(x=level, y=genes, color=Treatment))+
    facet_wrap(~Treatment*sex)+
    geom_point(size=2) + 
    geom_smooth(method="lm", color="black")+
    xlab("Methylation level") + 
    ylab("Gene expression")+
    ylim(0,2000)+
    #xlim(0,150)+
    theme_classic() + 
    theme(
      legend.position="right",
      axis.title=element_text(face="bold", size=14),
      axis.text=element_text(size=12, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=12)
      ); exon_plot

```
    
Plot by sample.  

```{r}
exon_plot2<-exon_file%>%
    ggplot(., aes(x=level, y=genes, color=Treatment, shape=sex))+
    facet_wrap(~sample_id)+
    geom_point(size=2) + 
    #geom_smooth(method="lm", color="black")+
    xlab("Methylation level") + 
    ylab("Gene expression")+
    ylim(0,2000)+
    #xlim(0,150)+
    theme_classic() + 
    theme(
      legend.position="right",
      axis.title=element_text(face="bold", size=14),
      axis.text=element_text(size=12, color="black"), 
      legend.title=element_text(face="bold", size=14), 
      legend.text=element_text(size=12)
      ); exon_plot2

```

    
    
    



